A Monte Carlo Synthetic Acceleration Method for the $SP_N$ Equations

### 2013-03-04 doc/spn/spn.tex
finished SPN derivation, on to spectral analysis

this as a matrix system:

@@

  -\nabla \cdot \mathbf{D} \nabla \mathbf{u} + \mathbf{A} \mathbf{u} =
  \mathbf{q}\:,

@@

This completes the derivation of the $SP_7$ equations for a single
energy group. As noted by Evans, lower order $SP_N$ approximations can
be generated by setting higher order even moments in this system to
zero (e.g. $\phi_6 = \phi_4 = 0$ yields the $SP_3$ equations).

@@

    +\\ \frac{1}{7\Sigma_3}\hat{\mathbf{n}}\cdot\nabla\Big(3\Big(\frac{1}{3}u_2
    - \frac{4}{15}u_3 + \frac{8}{35}u_4\Big) + 4\Big(\frac{1}{5}u_3 -
    \frac{6}{35}u_4\Big)\Big) + \frac{81}{128}\Big(\frac{1}{5}u_3
    -\\ \frac{6}{35}u_4\Big) - \frac{13}{64}\Big(\frac{1}{7}u_4\Big) =
    -\frac{1}{8}\phi_{b}\:,

@@

    +\\ \frac{1}{15\Sigma_7}\hat{\mathbf{n}}\cdot\nabla\Big(7\Big(\frac{1}{7}u_4\Big)\Big)

@@

  \hat{\mathbf{n}} \cdot \mathbf{D} \nabla \mathbf{u} + \mathbf{B}
  \mathbf{u} = \mathbf{s}\:,

@@

  -\nabla \cdot \mathbb{D} \nabla \mathbb{U} + \mathbb{A} \mathbb{U} =
  \mathbb{Q}\:,

@@

generated from the moment scattering matrices. Analogously, on the
boundaries we have:

@@

  \hat{\mathbf{n}} \cdot \mathbb{D} \nabla \mathbb{U} + \mathbb{B}
  \mathbb{U} = \mathbb{S}\:.

@@

%%---------------------------------------------------------------------------%%
\section{Spectral Analysis of the $SP_N$ Equations}
\label{sec:spectral_analysis}

>>>>

this as a matrix system of 4 equations in the case of the $SP_7$
approximation:

@@

  -\nabla \cdot D_n \nabla u_n + \sum_{m=1}^4 A_{nm} u_m =
  q_n\ \ \ \ \ \ \ n = 1,2,3,4\:,

@@

Note that the term $\sum_{m=1}^4 A_{nm} u_m$ in
Eq~(\ref{eq:spn_matrix}) couples the moments in each equation while
the diffusive term in each equation is only for a single
'psuedo-moment' $u_n$. This completes the derivation of the $SP_7$
equations for a single energy group. As noted by Evans, lower order
$SP_N$ approximations can be generated by setting higher order even
moments in this system to zero (e.g. $\phi_6 = \phi_4 = 0$ yields the
$SP_3$ equations).

@@

    +\\ \frac{1}{7\Sigma_3}\hat{\mathbf{n}}\cdot\nabla\Big
    (3\Big(\frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4\Big) +
    4\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big)\Big) +
    \frac{81}{128}\Big(\frac{1}{5}u_3 -\\ \frac{6}{35}u_4\Big) -
    \frac{13}{64}\Big(\frac{1}{7}u_4\Big) = -\frac{1}{8}\phi_{b}\:,

@@

    +\\ \frac{1}{15\Sigma_7}\hat{\mathbf{n}}
    \cdot\nabla\Big(7\Big(\frac{1}{7}u_4\Big)\Big) =
    -\frac{5}{128}\phi_{b}\:.

@@

  \mathbf{n} \cdot D_n \nabla u_n + \sum_{m=1}^4 B_{nm} u_m =
  s_n\ \ \ \ \ \ \ n = 1,2,3,4\:,

@@

  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
  \mathbb{A}_{nm} \mathbb{U}_m = \mathbb{Q}_n\:,

@@

generated from the moment scattering matrices:

@@

    (\mathbf{\Sigma}_0) &
    (-\frac{2}{3}\mathbf{\Sigma}_0) &
    (\frac{8}{15}\mathbf{\Sigma}_0) &
    (-\frac{16}{35}\mathbf{\Sigma}_0) \\
    (-\frac{2}{3}\mathbf{\Sigma}_0) &
    (\frac{4}{15}\mathbf{\Sigma}_0 + \frac{1}{3}\mathbf{\Sigma}_2) &
    (-\frac{16}{45}\mathbf{\Sigma}_0 - \frac{4}{9}\mathbf{\Sigma}_2) &
    (\frac{32}{105}\mathbf{\Sigma}_0 + \frac{8}{21}\mathbf{\Sigma}_2) \\
    (\frac{8}{15}\mathbf{\Sigma}_0) &
    (-\frac{16}{45}\mathbf{\Sigma}_0 - \frac{4}{9}\mathbf{\Sigma}_2) &
    (\frac{64}{225}\mathbf{\Sigma}_0 + \frac{16}{45}\mathbf{\Sigma}_2 + \frac{9}{25}\mathbf{\Sigma}_4) &
    (-\frac{128}{525}\mathbf{\Sigma}_0 - \frac{32}{105}\mathbf{\Sigma}_2 - \frac{54}{175}\mathbf{\Sigma}_4)
    (-\frac{16}{35}\mathbf{\Sigma}_0) &
    (\frac{32}{105}\mathbf{\Sigma}_0 + \frac{8}{21}\mathbf{\Sigma}_2) &
    (-\frac{128}{525}\mathbf{\Sigma}_0 - \frac{32}{105}\mathbf{\Sigma}_2 - \frac{54}{175}\mathbf{\Sigma}_4)
    (\frac{256}{1225}\mathbf{\Sigma}_0 + \frac{64}{245}\mathbf{\Sigma}_2 +
    \frac{324}{1225}\mathbf{\Sigma}_4 + \frac{13}{49}\mathbf{\Sigma}_6)
  \label{eq:A_block_matrix}
Analogously, for Marshak conditions on the boundaries we have:
  \hat{\mathbf{n}} \cdot \mathbb{D}_n \nabla \mathbb{U}_n +
  \sum_{m=1}^4 \mathbb{B}_{nm} \mathbb{U}_m = \mathbb{S}_n\:,

@@

with $\mathbb{S}_n$ the vector of group-wise boundary source term on
each boundary for each psuedo-moment and $\mathbb{B}_{nm}$ is an $N_g
\times N_g$ diagonal matrix with the value $B_{nm}$ on the
diagonal. For reflecting conditions, again we have $\nabla
\mathbb{U}_n = 0$ for all pseudo-moments.

\subsection{A Note on Spatial Discretization and Matrix Symmetry}
\label{subsec:spatial_discretization}
To this point, the formulation of the multigroup $SP_N$ equations
presented have discretized the transport equation in angle and energy
but have yet to consider the spatial component of phase space. We
don't explicitly consider it in this document but instead we will
briefly comment on possible means of discretization. Of primary
interest here is the discretization of the diffusive term in
Eq~(\ref{eq:spn_multigroup_system}) and the convective term on the
boundaries in Eq~(\ref{eq:spn_multigroup_bnd}). Many popular choices
for spatial discretization are available here and include finite
difference and finite element formulations. In the Denovo package at
ORNL, Evans has implemented a finite volume scheme for these
equations. Although arbitrary grids could be handled effectively
through a finite element scheme, the rectilinear grid used in Denovo
is easily discretized through the finite volume method in a
conservative form. This conservative form is ideal for the
diffusion-like equations in the domain given by
Eq~(\ref{eq:spn_multigroup_system}) and the effective boundary current
conditions given by Eq~(\ref{eq:spn_multigroup_bnd}) as the neutron
current is balanced from cell-to-cell, continuity of the flux is
preserved across cell/material boundaries, and boundary conditions are
naturally represented through cell-face currents. Furthermore, this
spatial discretization is a natural extension of the balance
principles used to arrive at the general transport equation. The
result of this finite volume discretization is one that is symmetric
for all equations in the domain with respect to the spatial components
of the solution.

Given the symmetry of the spatial discretization equations in the
domain given by monoenergetic Eq~(\ref{eq:spn_matrix}) form a
symmetric linear operator acting on the pseudo-moment vector to give:
\begin{equation}
  \mathbf{L}\mathbf{u}=\mathbf{Q}\:,
  \label{eq:matrix_system}
\end{equation}
where $\mathbf{L} = -\nabla \cdot \mathbf{D} \nabla +
\mathbf{A}$. Moving to the multigroup formulation in
Eq~(\ref{eq:spn_multigroup_system}) we have:
\begin{equation}
  \mathbb{L}\mathbb{U}=\mathbb{Q}\:,
  \label{eq:multigroup_matrix_system}
\end{equation}
where $\mathbb{L} = -\nabla \cdot \mathbb{D} \nabla +
\mathbb{A}$. Whether or not the resulting matrix is symmetric is
determined by the group scattering matrices, $\sigma_{sn}^{gg'}$, for
each of the moments. In general, the cross sections in these matrices
do not form a symmetric matrix and therefore the resulting linear
operator in Eq~(\ref{eq:multigroup_matrix_system}) will always be
asymmetric for multigroup problems. Solutions to this linear system
will then require techniques specifically formulated for asymmetric
systems.

%%---------------------------------------------------------------------------%%
\section{Spectral Analysis of the $SP_N$ Equations}
\label{sec:spectral_analysis}
Now that we have an understanding of the linear system generated by
the $SP_N$ approximation to the neutron transport equation and the
various parameters in the system that may be adjusted, we can consider
effective means of solution. As the solution is asymmetric, modern iterative
solvers should be considered for this task. The performance of modern
subspace methods are bound to various properties of the eigenvalue
spectrum of the linear operator while methods based on stationary
iterations have performance limits imposed by the spectral radius of
the sytem. In this section we will review the important spectral
properties to study for common iterative methods for asymmetric
systems. Based on these properties, we will then perform a
parameter-based spectral analysis for the linear operator generated by
the multigroup $SP_N$ equations.

\subsection{Iterative Solvers for Asymmetric Linear Systems}
\label{subsec:asymmetric_iterative_solvers}
The performance of different iterative solvers can depend on different
properties of the linear system. As a means to study these
differences, we use a 1992 paper by Nachtigal and colleagues
\cite{nachtigal_how_1992} the performs and in-depth analysis of 3
Krylov subspace methods that are still used today in modern physics
applications: conjugate gradient iteration for the normal equations
(CGN), generalized minimum residual iteration (GMRES), and a
biconjugate gradient-based methd (CGS).
